---
title: "You'll never run alone. Effect of 'cheering zones' on athlete performance in marathon races."
author:
  - name: Damien Dupré
    email: damien.dupre@ucd.ie
    affiliation: University College Dublin
    footnote: Corresponding Author
  - name: Aonghus Lawlor
    email: aonghus.lawlor@ucd.ie
    affiliation: University College Dublin
  - name: Barry Smyth
    email: barry.smyth@ucd.ie
    affiliation: University College Dublin
address:
  - code: University College Dublin
    address: The Insight Centre for Data Analytics, Belfield, Dublin 4

abstract: |

journal: "ISRE 2019"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
output: rticles::elsevier_article
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, message = FALSE, warning = FALSE, error = FALSE, echo = FALSE, fig.align="center")
library(papaja)
library(here)
# data wrangling
library(plyr)
library(data.table)
library(tidyverse)
# statistical modeling
library(geosphere) # claculate gps distance
# result plots
library(ggplot2)
```


```{r analysis_preferences}
set.seed(123)# Seed for random number generation
options(scipen = 999) #disable sci number format

# load data
dublin_2015_marathons_raw <- readRDS(here::here("marathon_cheering_data/dublin_2015_marathons_raw.rds")) %>%
  dplyr::filter(totaldistance > 40000) %>%
  dplyr::distinct(hashedathleteid, .keep_all = TRUE)
dublin_2015_marathons_raw$distance_conv <- as.array(
  lapply(
    strsplit(
      substr(dublin_2015_marathons_raw$distance,
             2,
             nchar(dublin_2015_marathons_raw$distance)-1),",",fixed=TRUE),
    as.numeric)
  )
dublin_2015_marathons_raw$time_conv <- as.array(
  lapply(
    strsplit(
      substr(dublin_2015_marathons_raw$time,
             2,
             nchar(dublin_2015_marathons_raw$time)-1),",",fixed=TRUE),
    as.numeric)
  )
```

```{r add-event-to-marathon-route}
# evaluate distance from marathon route
dublin_marathon_route <- read.csv(here::here("marathon_cheering_data/dublin_marathon_route.csv")) %>%
  dplyr::arrange(time) %>%
  dplyr::select(-time, -element)
#
distance_array <- NULL
for(i in 2:nrow(dublin_marathon_route)){
  DirectDistance <- as.numeric(geosphere::distm(c(dublin_marathon_route$lon[i-1],dublin_marathon_route$lat[i-1]),c(lon2=dublin_marathon_route$lon[i],lat2=dublin_marathon_route$lat[i]), fun = distHaversine))
  #To be more accurate, we also take into account difference in altitude between the points
  # AltitudeChange <- abs(z[i]-z[i-1])
  # if(AltitudeChange!=0){
  #   flight[i] <- sqrt(DirectDistance^2+AltitudeChange^2)
  # } else {
  #   flight[i] <- DirectDistance
  # } 
  distance_array <- c(distance_array,DirectDistance)
}
dublin_marathon_route <- dublin_marathon_route %>%
  dplyr::mutate(distance = c(0,distance_array))  %>%
  dplyr::mutate(distance = cumsum(distance)) %>%
  dplyr::distinct(distance,.keep_all = TRUE) %>%
  dplyr::filter(distance < 42195)
# attribuate new gps coordinates matching with marathon route points
dublin_marathon_event <- read.csv(here::here("marathon_cheering_data/dublin_marathon_event.csv")) %>%
  dplyr::select(-elevation, -time) %>%
  dplyr::filter(event=="cheering") %>%
  dplyr::mutate(event = as.character(event))
#
get_distance_event <- function(df){
  dublin_marathon_route %>%
    dplyr::mutate(abs_diff_lat = abs(lat - df[["lat"]])) %>% # Lat difference between Event and all the marathon route
    dplyr::mutate(abs_diff_lon = abs(lon - df[["lon"]])) %>% # Lon difference between Event and all the marathon route
    dplyr::mutate(closest_lat_lon = abs_diff_lat+abs_diff_lon) %>% # closest lat/lon from marathon route
    dplyr::top_n(-1,closest_lat_lon) %>% # filter closest lat/lon from marathon route (min_rank)
    dplyr::select(distance,lat,lon) %>%
    dplyr::mutate(event=df[["event"]])
}
distance_event <- dublin_marathon_event %>% 
  dplyr::rowwise() %>%
  dplyr::do(get_distance_event(.)) %>%
  as.data.frame() %>%
  dplyr::distinct(.keep_all = TRUE) # removes possible duplicates if cheering zones are too close (identical GPS coordinate from marathoon route)
```

```{r add-event-to-athlete-data}
list_athlete <- unique(dublin_2015_marathons_raw$hashedathleteid)
#
event_athlete <- plyr::ldply(list_athlete, function(athlete_id){
  #
  # print(athlete_id)
  #
  raw_data <- dublin_2015_marathons_raw %>%
    dplyr::filter(hashedathleteid == athlete_id)
  #
  long_data <- data.frame(
    hashedathleteid = raw_data$hashedathleteid,
    startlatapprox = raw_data$startlatapprox,
    startlngapprox = raw_data$startlngapprox,
    startdatelocal = raw_data$startdatelocal, 
    time = unlist(raw_data$time_conv), 
    distance = unlist(raw_data$distance_conv)
  )
  #
  long_data <- long_data %>%
    dplyr::mutate(timestamp = startdatelocal + time) %>%
    dplyr::mutate(pace_diff = c(NA,(1000.0/60)*diff(time)/diff(distance))) %>%
    dplyr::mutate(pace_diff = ifelse(pace_diff>25,25,pace_diff)) %>%  # remove artefact of pace too low
    dplyr::distinct(distance, .keep_all = TRUE) %>%
    dplyr::filter(distance < 42195 & distance != 0) %>%
    dplyr::select(distance, pace_diff)
  #
  # dplyr::mutate(pace_range = cut(pace_diff,
  #                                breaks = c(0,7.45,9.32,12.43,25),rigth=FALSE,
  #                                labels = c("Running", "Jogging","Brisk walking","Easy walking")))
  #
  distance_event_athlete <- distance_event %>%
    dplyr::filter(distance < max(long_data$distance)) %>%
    dplyr::filter(distance < 41700) # in order to evaluate a range of 1000m before and after without overlap
  #
  long_data <- long_data %>%
    dplyr::full_join(distance_event_athlete, by = "distance") %>%
    dplyr::arrange(distance)
  #
  long_data$pace_diff <- approx(long_data$distance,long_data$pace_diff, xout = long_data$distance)$y
  #
  long_data <- long_data %>%
    dplyr::select(-lon,-lat) %>%
    dplyr::mutate(hashedathleteid = athlete_id)
  #
  distance_event_range <- distance_event_athlete %>%
    dplyr::mutate(before = distance -1000) %>%
    dplyr::mutate(after = distance +1000)
  #
  before_cheering <- setDT(long_data)[distance %inrange% distance_event_range[, c("before","distance")]] %>%
    dplyr::filter(is.na(event)) %>%
    dplyr::mutate(event = "before")
  id <- match(long_data$distance, before_cheering$distance, nomatch = 0L)
  long_data$event[id != 0] <- before_cheering$event[id]
  #
  after_cheering <- setDT(long_data)[distance %inrange% distance_event_range[, c("distance","after")]] %>%
    dplyr::filter(is.na(event)) %>%
    dplyr::mutate(event = "after")
  id <- match(long_data$distance, after_cheering$distance, nomatch = 0L)
  long_data$event[id != 0] <- after_cheering$event[id]
  #
  long_data
})

event_athlete <- event_athlete %>%
  dplyr::mutate(is_cheering = ifelse(event == "cheering",1,0)) %>%
  dplyr::mutate(is_cheering = ifelse(is.na(is_cheering),0,is_cheering))%>%
  dplyr::mutate(is_cheering = as.factor(is_cheering)) %>%
  dplyr::mutate(before_after = case_when(
    event == "cheering" ~ 1,
    event == "after" ~ 1,
    event == "before" ~ -1,
    is.na(event) ~ 0
  )) %>%
  dplyr::mutate(before_after = as.factor(before_after))
```

```{r glm-is_cheering}
glmtest_ischeering <- glm(pace_diff~is_cheering*distance,data=event_athlete)
#papaja::apa_print(glmtest)
summary_glm_ischeering <- summary(glmtest_ischeering) %>%
  magrittr::use_series(coefficients)

glmtest_before_after <- glm(pace_diff~before_after*distance,data=event_athlete)
#papaja::apa_print(glmtest)
summary_glm_before_after <- summary(glmtest_before_after) %>%
  magrittr::use_series(coefficients)
```

```{r cheering-gain}
mean_group <- event_athlete %>%
  dplyr::filter(pace_diff !=0) %>%
  dplyr::group_by(is_cheering, hashedathleteid) %>%
  dplyr::summarise(m = mean(pace_diff)) %>%
  tidyr::spread(is_cheering, m) %>%
  dplyr::mutate(diff_cheering = `0`-`1`) %>%
  dplyr::mutate(effect_cheering = (diff_cheering/`0`))
```

Although the emotion literature advocates for an influence of positive emotions on sports performance [@mccarthy2011positive; @vast2010emotions], quantifying this influence remains a challenge. Among remarkable athletic performances, marathon races are a relevant example of this influence. The establishment of "cheering zones" during marathon races shows how positive emotions and social support are important for athletes to enhance their performances [@buman2008experiences]. Our aim is to quantify the behavioural impact of these cheering zones on athletes pace during marathon races.

```{r}
#Even if studies have identified the role of positive emotion on athletes' perfomance from self-report [@shipway2013organisations], our aim is to quantify the behavioural impact of these cheering zones on athletes pace during marathon races.

#the Strava app gives feedback to athletes about race distance, duration and elevation in real time. From distance and duration time series,

#Our results are supporting this idea rather than potential effect of social presence [@lombard1997heart] which would have desappeared once the athletes are out of the cheering zones [@morgado2011close].
```

In collaboration with Strava Inc. (athlete monitoring application) we analysed the data of `r nrow(mean_group)` athletes who have finished the Dublin marathon in 2015. By analyzing their GPS information the Strava app is calculating the evolution of athletes' pace (min/km) during the marathon. We compared athletes' pace before, during and after the cheering zones in order to identify the influence of positive emotion and social support on athletes' performance.
  
Generalized Linear Models show not only an effect of cheering zones on atheltes' pace (*t* = `r round(summary_glm_ischeering[2,3],2)`, *p* <`r format.pval(summary_glm_ischeering[2,4], eps = 0.001,digits = 1)`) but also an effect of the localisation of these cheering zones (*t* = `r round(summary_glm_ischeering[4,3],2)`, *p* `r format.pval(summary_glm_ischeering[4,4], eps = 0.001,digits = 1)`). Athletes tend to increase their pace by `r scales::percent(mean(mean_group$effect_cheering))` after each cheering zones on average but this effect tend to decrease along the marathon race. This last result is supported by the comparison athlete's pace comparison before and after the cheering zones which is significant only by taken into account their localisation (*t* = `r round(summary_glm_before_after[6,3],2)`, *p* `r format.pval(summary_glm_before_after[6,4], eps = 0.001,digits = 1)`). 

Our results are supporting the theory of individual zones of optimal functioning (IZOF) for which feeling the support of others in cheering zones would helps athletes to find the motivation to sublim their performance [@hagtvet2007consistency]. 

References {#references .unnumbered}
==========